### 2023-02-12 article.tex
Review

The experiment described above is run by code. 
It was code that detected the relationship between the genotype
(in this case, SNP \verb|rs12133641|) 
and the phenotype (in this case, the concentration of IL6RA).
To be more precise, it was code that read the data,
subsetted the data, removed outliers, 
performed the statistics and generated the plots.
For the rest of the discussion, we assume that the
code is available to us (note that it is not
published with the paper, see section \ref{sec:code-availability} 
for a glum estimate of the chance of obtaining the code).

@@

The reason behind paradata is, among others,
to assure reproducible research \cite{huvila2022improving}.
Reproducibility (i.e. to reproduce the same results) 
and replicability (i.e. to reconclude a conclusion)
are fundamental characteristics of scientific studies \cite{patil2019visual}.
For computational science, it seems to be relatively easy to 
reproduce an experiment, as all it takes is a computer, electricity,
an optional internet connection, the code and the data.
In practice, however, it appears that 
the academic culture to reproduce results 
has been lost over time in general \cite{peng2011reproducible},
with labs that embrace reproducibility (for example, \cite{barba2016hard})
are the exception.
One suggested way forward is to make the reproduction of 
research a minimal requirement for publication \cite{peng2011reproducible}.

@@

From a broader perspective, code should be published, as it holds the
ground truth of an experiment; it does the actual work.
The more complex the computation pipeline is, the easier it is
to have a mismatch between the article (that describes what the
code does) and the code (that actually does the work).
It is easy (tempting?) to overlook how easy it is to have a mismatch
between paper and code.

@@

It is the paper that accompanies the code,
as it is the code that generates the results.
When humans are fallible and code gets bigger, the likelihood of
a mismatch between the English paper and the code increases.
We know that the most common errors are 
simple (and may legitimately fear that those simple errors are common) 
\cite{baggerly2009deriving},
and this example is not too far off from the
examples given in that paper.

@@

Preserving code in a reliable way is rare \cite{barnes2010publish}.
In general, within software development, 
changes at all levels are common
and instead of ignoring that, it should be embraced \cite{beck2000extreme}.
This section discusses the preservation of code for a short, medium
and long term.

@@

To preserve code for a short term, 
a code hosting website is a good first step.
A code hosting website is a website where 
its users can create dedicated pages (called 'repositories')
for a project, upload code and interact with that code.
There are multiple code hosting websites, 
with GitHub being the most popular one \cite{cosentino2017systematic}.
The \LaTeX code of this chapter is hosted on GitHub 
as well, at \url{https://github.com/richelbilderbeek/chapter_paradata}.
The use of code hosting websites
has increased strongly \cite{russell2018large},
accommodates collaboration \cite{perez2016ten}
and improves transparency \cite{gorgolewski2016practical}.
See \cite{cosentino2017systematic} for an extensive overview of
research conducted on GitHub.

@@

However, for code to work longer, one can put the code and its dependencies in
a container, that should be best in standing the test of time.

@@

The most reproducible way of submitting the code of an experiment,
is by providing the code with all the (software) dependencies 
it needs installed in a virtual container.
This is close to the golden standard suggested by 
\cite{peng2011reproducible} 
\iffalse
(see also Figure \ref{fig:peng2011reproducible})
\fi
,
however, that article was written before virtualization existed.
Virtualization allows one to provide a standard environment,
that can be used on any other computer.
The best known examples of container software are Docker and Singularity.
Due to access rights, Singularity is the only viable option to run
an experiment on a 
high performance computing (HPC) environment (e.g. a computer cluster).

@@

For code to be ideal paradata, it must be usefully 
documented \cite{huvila2022improving}.
What is 'useful' is determined by its intended use: 
to reproduce, replicate or extend \ref{sec:making-code-useful-paradata}.
The amount of documentation is, however, also determined
by the recommended practices for writing code in general.

For purposes of reproducing (same code, same data) code,
it should at least be documented how the code is run.
However, 57\% out of 56 Science magazine corresponding authors 
that shared their code has documented their code well enough to run it 
and to be understood what it ought to do \cite{stodden2018empirical}.

@@

For code to be ideal paradata, 
its creation and documentation need to align, 
\emph{not least because they shape each other} \cite{huvila2022improving}.
This emphasized part of the quote resonates strongly with the idea of 
literate programming \cite{knuth1984literate},
in which documentation and code is developed hand-in-hand.
It is the practice of writing a document mixed with code that 
will be run.
Contemporary examples of this idea are, among others, 
vignettes \cite{wickham2015r} (for the R programming language \cite{r}) 
and Jupyter notebooks \cite{wang2020assessing} 
(for the Julia \cite{Julia-2017}, Python \cite{van1995python} and R \cite{r}
programming languages).
rOpenSci, a community that, among others, reviews programming
code (\cite{ram2013ropensci,ram2018community}), is 
one example of where literate programming is enforced:
code must have extensive runnable examples
as part of the documentation \cite{ropensci_2021_6619350}.
Developing code and documentation at the same time 
leads to both better code and documentation \cite{reenskaug1989environment}.
In general, when developing software, 
it is recommended to 
to write documentation while writing software,
as well as to include many examples \cite{lee2018ten}.

@@

Code must be more extensive when it is (intended to be) 
used on different data,
as then 'code must act as a teacher for future developers' \cite{sadowski2018modern}.
Error handling is one of the mechanisms to do so.
A simple example is when one (e.g. a BSc student) 
needs to calculate the variance of a distribution.
Commonly one needs to call a function to do so, 
e.g. \verb|var| in R, that needs the values of the distribution
as input.
However, to calculate the variance of a distribution, at least two values
are required (else a division by zero occurs in the calcation of the variance). 
When the variance of (zero or) only one value is requested,
an error message can helpfully indicate that the function 
needs at least two values to work as expected.

@@

The percentage of (lines of) code tested is called the code coverage.
Code coverage correlates with code quality \cite{horgan1994,del1995correlation}. 
The non-profit organisation rOpenSci \cite{ram2018community},
which peer-reviews code,
has made it a prerequisite to have a code coverage of 100\%.
Also here, CI opens up the possibility to informally annotate the
code coverage. 
For example, build badges
\iffalse
(e.g. as shown in Figure \ref{fig:badge_codecov})
\fi
are visual indications on a website that can, among others,
show a certain R package's code has a 100\% code coverage.

Hence, code coverage is important metadata, that can automatically
be acquired when using code hosting website.

@@

Code hosting websites such as GitHub help developers to follow
best practices. One of these is to add a software license \cite{jimenez2017four},
so that it is clear that the software can be reused.
About two-thirds of 56 Science magazine corresponding authors 
that shared their code supply a software license \cite{stodden2018empirical}.

@@

More than half of all scientists have their code 
rarely or never reviewed \cite{vable2021code},
although code reviews are known to
accelerate learning of the developers,
improve the quality of the code
and resulting to an experiment that is likelier to be reproducible
\cite{vable2021code}.
Within software development best-practices, 
code reviews are recommended as well \cite{wilson2014best}

@@

There are multiple ways of publishing code,
that require increasingly more knowledge to do and result
in increasingly more reproducible experiments:
(1) publishing the code as text only, e.g. by printing
it in a Supplementary Materials, (2) publishing
the code on a hosted website, e.g on GitHub, with
added features (3) publishing 
the installed code within a virtual environment,
for example, using a Singularity container.
In this chapter, details of each of these ways will be described,
including advantages, disadvantages, as well as how to
make the code more FAIR data \cite{wilkinson2016fair}.

@@

Ideally, to re-do an experiment, one would need a one-liner to a container,
for example, \verb|singularity run my_experiment.sif run|.

@@

Containers allow a computation experiment to be highly reproducible:
given the same data, an experiment put into a container will give
the same results on different platforms, at least in theory.
In practice, differences may be observed when peripheral factors
are different, such as the random numbers as generated by an operating
system, or data that are downloaded from additional online sources.

@@

% \paragraph{Hosted code has important metadata}

Hosted code has many additional features, such as
a version-controlled history, continuous integration scripts
and automatically generated metrics.
A code hosting website produces and shares these metadata.
These metadata can help to determine the
reproducibility, correctness and quality of code
of computational experiments.
Below, those features and why these are important are discussed.

% \paragraph{Suggestions for knowledge management}

@@

Code used in academic research complies badly with FAIR principles,
where unpublished code is the worst offender.
However, any idealistic researcher does not have the
tools to following FAIR principles in an exemplary way:
there are no standards in how a scholarly article
refers to academic code (i.e. in a machine-friendly way).
Creating a (Singularity) container is the best a researcher
can do to ensure his/her code is reusable,
yet, there are no incentives to make code accessible, interoperable
or re-usable, with standarized metadata attributes being absent.

@@

@@

For research to be reproducible, one ideally has access to
both the data used and the code.
In some fields, such as genetic epidemiology, the data are
sensitive, hence cannot be released,
yet there are methods being devised to run code on sensitive
data with assured privacy \cite{zhang2016review,azencott2018machine}.
Additionally, it follows software development's best practices 
(especially the practice to test code automatically) 
to release a simulated/public dataset, with
the additional benefit that this dataset can be used for comparisons.

@@

Code needs maintanance:
the libraries that are used by an experiment
will change over time, inevitably causing the experiment to fail
over time. 
The degradation of software is a known feature for nearly 
four decades (see the definition of 'bit rot' at \cite{steele1983hacker}),
here a more modern (and in this context superior)
definition is used: software collapse \cite{hinsen2019dealing},
in which software fails due to dependencies on other 
software.
Where that paper recommends using older software foundations 
with multiple implementations \cite{hinsen2019dealing},
this paper recommends a modern solution that potentially has more 
longevity:
to publish the code installed in a virtual environment
(including the dependencies used at that time), as described below.
Write code that will stand the test of time perfectly
is deemed 'impossible' \cite{benureau2018re}.

@@

However, upon the find of a new bug, the question is:
does all the research that use that software still 
result in the same conclusions?
To prevent authors chosing to hide bug reports, 
a code hosting website usually has a public place to report bugs,
to increase transparency in the handling of bug reports.

@@

% \paragraph{The drawbacks of publishing version-controlled code}

Is it worth it to publish version-controlled code?
For an author, 
there is additional training involved, and also here,
publishing code alongside an experiment opens up
the possibily to receive questions regarding that code,
as well as other interactions, including helpful contributions.

% \paragraph{The drawbacks of publishing a running version of code}

Is it worth it to archive running versions of code?
For an author, 
there is additional training involved.
There is no FAIR infrastructure for Singularity containers.

% % \paragraph{Still need the same data to do a reproducible experiment}

% % \paragraph{Recommendations}

Being able to re-do an experiment is a core principle of the scientific method.
Publishing only a paper about a computational experiment is not enough,
as the results are too likely to mismatch that English description.
Journal should make it mandatory for authors
to publish their code alongside a computational experiment.
Authors should follow the FAIR principes for their code as well,
as can be done using the infrastructure as supplied by, 
for example, GitHub and GitLab, opening up new angles in
metascience regarding computational experiments.
Knowledge managers should create the infrastructure for the preservation
of runnable experiments, to allow scientist to upload a Singularity
container, so these are as well following the FAIR principles.

@@

This paper started with some suggestions to 
make paradata useful for data re(use):
paradata should be comprehensive, comprehensively documented
and friendly to computers \cite{huvila2022improving}.
When applying these general recomendations to code, 
this list can be phrased more precisely:
code should be comprehensive in supplying 
automatically generated metadata (such as commit history and code coverage).
The documentation should be as extensive as recommended by the 
software development literature. To be computer friendly,
code should be able to run on at least one independent entity,
such as a GitHub Actions server or from a Singularity container.

@@

When research truly needs to be reproducible, putting the code 
of an experiment into a container is a good solution, as it 
can stand the test of time better.
Creating such a container, however, requires more skill
that -as of today- is not rewarded,
partly as containers follow FAIR data principles only poorly.
Runnable code is the pinnacle of reproducible research.

@@

The simplest and most impactful step to make code more useful paradata
is, however, to publish it on a code hosting website 
along a publication. From then on, the next steps can be taken 
gradually as the skills of the author(s) progress.

>>>>

The experiment described above is run by code. 
It was code that detected the relationship between the genotype
(in this case, SNP \verb|rs12133641|) 
and the phenotype (in this case, the concentration of IL6RA).
To be more precise, it was code that read the data,
subsetted the data, removed outliers, 
performed the statistics and generated the plots.
For the rest of the discussion, we assume that the
code is available to us (if not, see section \ref{sec:code-availability} 
for a glum estimate of the chance of obtaining the code).

% \paragraph{Why useful code}

There are multiple reasons why (useful) code matters,
and these are the same reasons as why useful paradata matters:
code gives an understanding how the raw results 
and the subsequent scholarly knowledge is obtained from an experiment.
Additionally, code helps researchers from different fields 
to understand each other and collaborate.
Additionally, code helps Open Science reach its goals of openness and
transparency.
The core of these reasons it to achieve reproducible science:
that any person in any field can redo a computational experiment
and see exactly what happened.

@@

For computational science, it may appear to be relatively easy to 
reproduce an experiment, as all it takes is a computer, electricity,
an optional internet connection, the code and the data.
In practice, however, 
only 18\% of 180 studies are easily reproducible \cite{stodden2018empirical}.
To some, it appears that 
the academic culture to reproduce results 
has been lost over time \cite{peng2011reproducible},
with labs that embrace reproducibility (for example, \cite{barba2016hard})
are the exception.
One suggested way forward is to make the reproduction of 
research a minimal requirement for publication \cite{peng2011reproducible}.

@@

Code holds the ground truth of an experiment; it does the actual work.
The more complex the computation pipeline is, the easier it is
to have a mismatch between the article (that describes what the
code does) and the code (that actually does the work).
It is easy (tempting?) to overlook how easy it is to have a mismatch
between paper and code.

@@

It is the paper that accompanies the code,
as it is the code that generates the results.
When humans are fallible and code gets bigger, the likelihood of
a mismatch between the English paper and the code increases.
We know that the most common errors are 
simple (and may legitimately fear that those simple errors are common) 
\cite{baggerly2009deriving, vable2021code},
and this example is not too far off from the
examples given in that paper.

@@

Preserving code in a reliable way is rare \cite{barnes2010publish}.
In general, within software development, 
changes at all levels are common
and embracing those changes is recommended \cite{beck2000extreme}
and useful code does so.
This section discusses the preservation of code for a short, medium
and long term.

@@

To preserve code for a short term, 
a code hosting website is a good first step.
A code hosting website is a website where 
its users can create dedicated pages (called 'repositories')
for a project, upload code and interact with that code.
There are multiple code hosting websites, 
with GitHub being the most popular one \cite{cosentino2017systematic}.
The \LaTeX code of this chapter is hosted on GitHub 
as well, at \url{https://github.com/richelbilderbeek/chapter_paradata}.
The use of code hosting websites
has increased strongly \cite{russell2018large},
accommodates collaboration \cite{perez2016ten}
and improves transparency \cite{gorgolewski2016practical},
due to its inherent computer-friendliness.
See \cite{cosentino2017systematic} for an extensive overview of
research conducted on GitHub.

@@

However, for code to work longer, one can put the code and its dependencies in
a container, that should be best in standing the test of time,
as the degradation of software is a known feature for nearly 
four decades (see the definition of 'bit rot' at \cite{steele1983hacker}),
here a more modern (and in this context superior)
definition is used: software collapse \cite{hinsen2019dealing},
in which software fails due to dependencies on other 
software.
Where that paper recommends using older software foundations 
with multiple implementations \cite{hinsen2019dealing},
this paper recommends a modern solution that potentially has more 
longevity:
to publish the code installed in a virtual environment
(including the dependencies used at that time), as described below.

@@

The most reproducible way of submitting the code of an experiment,
is by providing the code with all the (software) dependencies 
it needs installed in a virtual container.
This is close to the golden standard suggested by 
\cite{peng2011reproducible} 
\iffalse
(see also Figure \ref{fig:peng2011reproducible})
\fi
,
however, that article was written before virtualization existed.
Virtualization allows one to provide a standard environment,
that can be used on any other computer.
The best known examples of container software are Docker and Singularity.
Due to access rights, Singularity is the only viable option to run
an experiment on a 
high performance computing (HPC) environment (e.g. a computer cluster),
as are commonly needed in the field of genetic epidemiology experiments.

@@

For code to be ideal paradata, it must be usefully 
documented \cite{huvila2022improving}.
What is 'useful' is determined by its intended use: 
to reproduce, replicate or extend (see section \ref{sec:making-code-useful-paradata}).
The amount of documentation is, however, also determined
by the recommended practices for writing code in general.

For purposes of reproducing (same code, same data) code,
it should at least be documented how the code is run.
However, 57\% out of 56 Science papers with obtainable code (in 
total there were 180 papers) have documented their code well enough to run it 
and to be understood what it ought to do \cite{stodden2018empirical}.

@@

For code to be ideal paradata, 
its creation and documentation need to align, 
\emph{not least because they shape each other} \cite{huvila2022improving}.
This emphasized part of the quote resonates strongly with the idea of 
literate programming \cite{knuth1984literate},
in which documentation and code is developed hand-in-hand.
Literate programming is the practice of writing a document mixed with code that 
will be run.
Contemporary examples of this idea are, among others, 
vignettes \cite{wickham2015r} (for the R programming language \cite{r}) 
and Jupyter notebooks \cite{wang2020assessing} 
(for the Julia \cite{Julia-2017}, Python \cite{van1995python} and R \cite{r}
programming languages).
rOpenSci, a community that, among others, reviews programming
code (\cite{ram2013ropensci,ram2018community}), is 
one example of where literate programming is enforced:
code must have extensive runnable examples
as part of the documentation \cite{ropensci_2021_6619350}.
Developing code and documentation at the same time 
leads to both better code and documentation \cite{reenskaug1989environment}.
In general, when developing software, 
it is recommended to 
to write documentation while writing software,
as well as to include many examples \cite{lee2018ten}.

@@

Code must be more extensive when it is (intended to be) 
used on different data,
as then 'code must act as a teacher for future developers' \cite{sadowski2018modern}.
Error handling is one of the mechanisms to do so.
One example is, again, when doing a t-test:
to compare the means of two distributions, each distribution
must have at least two values
and, if not, the programming language should signal the error clearly.
In genetic epidemiology, it is common to have incomplete or missing data,
so analyses should take into account that there are not always enough
observations to do a t-test.
The analytical code should clearly signal this problem.

@@

The percentage of (lines of) code tested is called the code coverage.
Code coverage correlates with code quality \cite{horgan1994,del1995correlation}. 
The non-profit organisation rOpenSci \cite{ram2018community},
which peer-reviews code,
has made it a prerequisite to have a code coverage of 100\%.
CI opens up the possibility to informally annotate the
code coverage. 
For example, build badges
\iffalse
(e.g. as shown in Figure \ref{fig:badge_codecov})
\fi
are visual indications on a website that can, among others,
show a certain R package's code has a 100\% code coverage.

@@

It is considered good practice
to add a software license \cite{jimenez2017four},
so that it is clear that the software can be reused.
About two-thirds of 56 Science magazine corresponding authors 
that shared their code supply a software license \cite{stodden2018empirical}.

@@

In an informal Twitter poll,
more than half of all 315 responding scientists 
have their code rarely or never reviewed \cite{vable2021code},
although code reviews are known to
accelerate learning of the developers,
improve the quality of the code
and resulting to an experiment that is likelier to be reproducible
\cite{vable2021code}.
Within software development best-practices, 
code reviews are recommended as well \cite{wilson2014best} 
The observation that code is reviewed, makes it likelier to be correct.

@@

@@

@@

Containers allow a computation experiment to be highly reproducible:
given the same data, an experiment put into a container will give
the same results on different platforms, at least in theory.
In practice, differences may be observed when peripheral factors
are different, such as the random numbers as generated by an operating
system, or data that are downloaded from online (and hence, changing) sources.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Suggestions for knowledge management}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Making code FAIR.

\subsection{Hosted code}

\subsection{CI code}

\subsection{Code in container}

@@

@@

For research to be reproducible, one ideally has access to
both the data used and the code.
In some fields, such as genetic epidemiology, the data are
sensitive, hence cannot be released,
yet there are methods being devised to run code on sensitive
data with assured privacy \cite{zhang2016review,azencott2018machine}.

@@

@@

Code needs maintanance:
the libraries that are used by an experiment
will change over time, inevitably causing the experiment to fail
over time. 
Write code that will stand the test of time perfectly
is deemed 'impossible' \cite{benureau2018re}.

@@

However, upon the find of a new bug, the question is:
does all the research that use that software still 
result in the same conclusions?
To prevent authors chosing to hide bug reports, 
a code hosting website usually has a public place to report bugs,
to increase transparency in the handling of bug reports.
Should there be a retraction notice on code?

@@

@@

This paper started with some suggestions to 
make paradata useful for data re(use):
paradata should be extensive, comprehensively documented,
with the creation of documentation and code going hand-in-hand,
as well as friendly to computers \cite{huvila2022improving}.

Before applying these features, the first step is to publish 
the code.

When applying these general recomendations to code, 
this list can be phrased more precisely:
code should be comprehensive in supplying 
automatically generated metadata (such as commit history and code coverage).
The documentation should be as extensive as recommended by the 
software development literature and have co-evolved with the
code following the best practices in literate programming. 
To be computer friendly,
code should be able to run on at least one independent entity,
such as a GitHub Actions server or from a Singularity container.

The second step for code to be useful paradata is to add CI, 
to stand the test of time better.
The ultimate step is to publish code in a container.

@@

When research truly needs to be reproducible, putting the code 
of an experiment into a container is a good solution, as it 
can stand the test of time better.
Creating such a container, however, requires more skill
that -as of today- is not rewarded,
partly as containers follow FAIR data principles only poorly.
However, such runnable code is the current pinnacle of reproducible research.

@@

The simplest and most impactful step to make code more useful paradata
is, however, to publish it on a code hosting website 
along a publication. From then on, the next steps can be taken 
gradually as the skills of the author(s) progress.
To quote \cite{barnes2010publish}: Publish your code, it is good enough.
